# Extensible Transmission Dynamics 

If we want to use mathematical models to evaluate policies, we must acknowledge the limitations of the Ross-Macdonald model. Its simplicity is useful as a concise way of understanding malaria transmsission through the parasite's life-cycle. There is, however, a great deal of complexity that the Ross-Macdonald model does can not address, and some of it matters a great deal in policy. We must extend the model using new mathematical constructs to deal with several features of malaria: 

+ malaria immunity and disease in cohorts of humans as they age; 

+ anti-malaria drugs and other pharmaceutical interventions; 

+ heterogeneous exposure

+ spatial dynamics and malaria importation; 

+ temporal dynamics including seasonality, perhaps exogenously forced by weather; 

+ mosquito ecology and behavior; 

+ and vector control. 

In the following chapters, we will introduce and critically review the models that have been developed to address these topics in some detail. As we take on these new features and add complexity, we will (in effect) be going around the parasite and mosquito life-cycles over and over again, each time pursuing a new theme. 

We want to pursue the idea of scaling complexity in dynamic models ranging from simple and abstract to highly realistic. We want to arrive at a happy compromise -- a model that is (as Einstein probably never exactly said [@RobinsonA2018DidEinstein]) "as simple as possible, but no simpler." It might be hard to convince anyone that a model of intermediate complexity is good enough unless at least _some_ of the models have clearly gone too far. We don't want to add new features to our models unless they would affect some important aspect of malaria transmission dynamics and control. Fortunately, we can pursue many of these questions about what details are biologically relevant through model-model comparison. We think of all these models as crude approximations to a messy reality -- the question we want to address is whether the approximations are good enough to use as a basis for giving policy advice. Building very complex models that are also mathematically and biologically sound and meaningful is a daunting task, but we have developed a modular system for building models that streamlines the model building process. 

To get started, we will reintroduce the Ross-Macdonald model in a form that is easier to extend than the one Macdonald made famous [@SmithDL2012_RossMacdonald]. We will start with a model that first appeared in  a 1982 book chapter written by Joan Aron and Robert May [@AronJL1982PopulationDynamics]. It is a good starting point because it is *extensible.* 

We chose it because the variables represent population densities, which are used to compute proportions, like *prevalence.* In the version of the Ross-Macdonald we presented before (in [Rebuilding Macdonald's Model]), the variables are proportions. In some models, we would like the *total* number of hosts to change over space or time, but if our state variables are proportions, then these densities appear in the denominators, which would make them much more difficult to modify. By writing down equations using densities, it is far easier to add complexity. 

The model is extensible because, as we will see, mosquito population density is a variable, not a term. A parameter in the equation describing changes in mosquito density provides a link to aquatic mosquito ecology, which makes it possible to develop models for mosquito population dynamics, when that is an important consideration. In this version, we find that the formulas describing $R_0$ and vectorial capacity tell the story of parasite transmission through two blood meals. We will thus update the classical formulas. While Macdonald's analysis and formulas are familiar to many, they were incomplete [@SmithDL2004_Statics; @SmithDL2021_NewTestOldMosquitoes]. We develop a formula for vectorial capacity that is consistent with the intent of the original, but our analysis of sensitivity to parameters includes effects on mosquito ecology [@BradyOJ2015AdultVector]. (A lengthy and philosophical discussion of the history and its failings is planned.) 

We chose it because it is *realistic.* Time does not appear in most versions of the Ross-Macdonald model: the equations are *autonomous*. These equations use time to drive a seasonal pattern: they are *non-autonomous*. Since we know we are interested in dealing with exogenous forcing, we start out with a model that is forced.  

While the following model is basic, we recommend reading it, if only because we introduce concepts and conventions that are important for the software design.


## The Aron-May Model 

We define the following variables: 

+ $M$ is the density of mosquitoes.

+ $Y$ is the density of infected mosquitoes. 

+ $Z$ is the density of infectious mosquitoes. 

+ $X$ is the density of infected humans. 

In dynamical systems, we ask how the variables (*i.e.* $M$, $Y$, $Z$, and $X$) change over time. In the following, we describe the changes on variable about a time. 


### Mosquito Ecology 

For our first equation, we start with adult, female mosquito populations. (It is tiresome to repeat *adult, female* each time, and we're ignoring male mosquitoes at this point anyway, so *mosquito* hereafter means *adult, female mosquito*, unless we say otherwise.) The number of mosquitoes is changing as new adults emerge from aquatic habitats or die. 

To model changes in $M$, we assume the following: 

+ mosquitoes emerge from aquatic habitats at the rate of $\Lambda(t)$ adults, per day; 

+ mosquitoes die at a constant rate, $g$. This is equivalent to assuming that the mosquito lifespan is exponentially distributed with a mean $1/g$. The fraction surviving one day is $e^{-g}$. 

Our first equation describes changes in the number of mosquitoes: 

\begin{equation}
\frac{dM}{dt} = \Lambda(t) - g M
\end{equation}


### Blood Fed Mosquitoes

At this point, we will take a detour and define a variable describing the density of mosquitoes that have blood fed at least once, $V$. After blood feeding, a mosquito is either gravid or *parous*, meaning its ovaries are distended from laying an egg batch. We do this, in part, because the fraction of mosquitoes that are parous is routinely collected, and because it gives us a chance to focus on blood feeding.

To describe *blood feeding*, we assume the following:

+ mosquitoes blood feed at the rate $f$, per mosquito, per day; in this model, this implies that the waiting time to a blood meal is $1/f$ days.

+ a fraction of all mosquito blood meals, $q$, is taken on humans; we call this the *human fraction*

+ the human blood feeding rate is the product of these two parameters, $fq$, which is defined as the number of human blood meals, per mosquito, per day. 

The number of human blood meals by a population of vector mosquitoes, per person, per day is called the human biting rate (HBR). In this model, HBR is given by a formula:

$$\mbox{HBR} = \frac{fqM}{H}$$ 

Later, we discuss the correspondence between the HBR in models and data.


\begin{equation}
\frac{dV}{dt} = f q (M-V) - g V
\end{equation}

We won't use $V$ to describe the dynamics of infection, but we might find it useful to understand how parity changes in mosquito populations. 

### Infected Mosquitoes

Mosquitoes become infected after blood feeding on an infectious human. To model changes in $Y$, we extend the model of blood feeding to include infection. We need to know what fraction of blood meals end up infecting a mosquito that has not already been infected. 

To model changes in $Y$, we need to describe *infection rates*. We assume the following: 

+ a fraction of human blood meals, infects mosquitoes. We call this quantity *net infectiousness* (NI) and (for reasons that we will discuss in a moment), we give it a name, $\kappa$: 

\begin{equation}
\kappa(t) = c \frac{X(t)}{H}
(\#eq:kappaDef)
\end{equation}

+ infected mosquitoes die at the same rate as uninfected mosquitoes.  


We can now write down our second equation describing changes in the number of infected mosquitoes: 

\begin{equation}
\frac{dY}{dt} = f q \kappa (M-Y) -g Y
\end{equation}


### Infectious Mosquitoes

To become infectious, a mosquito has to become infected and then survive through the extrinsic incubation period (EIP). We assume:

+ mosquitoes become infectious after a fixed delay, $\tau$ days, called the EIP. The fraction of mosquitoes that survive through the EIP is $e^{-g \tau}$.

+ infectious mosquitoes die at the same rate as other mosquitoes.

For a mosquito to become infectious, it must have become infected $\tau$ days ago and survived through $\tau$ days with probability $e^{-g\tau}$. To write this in equations, we use a subscripted $\tau$ to denote the value of a variable ($M$, $Y$ or $X$) or term ($\kappa$) at time $t-\tau$. For example $X_\tau$ is the number of people who were infected and infecious at time $t-\tau$, and $M_\tau$ is the number of mosquitoes at time $t-\tau$. 

The number of infectious mosquitoes that are added to the population at a point in time includes all the mosquitoes that became infected at time $t-\tau$ and survived the EIP. This is our third equation describing changes in the number of infectious mosquitoes: 

\begin{equation}
\frac{dZ}{dt} = f q  \kappa_\tau (M_\tau-Y_\tau) e^{-g\tau} -g Z
\end{equation}

Here, $Z$ represents the number of mosquitoes with *sporozoites* in their salivary glands. The *fraction* of mosquitoes with sporozoites in their salivary glands has been called the *sporozoite rate* (SR), which in our notation is 

$$ z = \frac{Z}{M}$$ 

The number of bites by vector mosquitoes, per person, per day is called the entomological inoculation rate (EIR). It is defined as the product of the HBR and the SR: 

$$\mbox{EIR} = \mbox{SR} \times \mbox{HBR}$$

In our notation, the EIR is: 

$$\mbox{EIR} = z \frac{fqM}{H} = \frac{fqZ}{H}$$ 
As with the HBR, we would like to know how to connect estiamted values of the EIR to our formulas. Since that's *really* complicated, we've spent a lot of time in the following sections discussing it. 


### Infected Humans {#RossEqn}

Humans become infected after being bitten by an infectious mosquito. We assume the following: 

+ A fraction $b$ of all bites by infectious mosquitoes cause an infection. 

+ The hazard rate for infection, also called the *force of infection* (FoI) and denoted $h$ is $b \times$ EIR: $$h = fqb \frac{Z}{H}$$

+ Infections clear at the rate $r$, per infection, per day (the average time to clear is $1/r$ days), and after clearing an infection a person becomes susceptible to infection again. 

We can now write down our fourth equation describing changes in the number of infected humans: 

\begin{equation}
\frac{dX}{dt} = h (H-X) - r X 
\end{equation}

### ...as a System 


While we presented these equations one at a time, they work as a system. To see it all at once, we write it here as a system with four equations and two terms: 

\begin{equation}
\begin{array}{rl}
\frac{dM}{dt} &= \Lambda(t) - g M \\
\frac{dY}{dt} &= fq\kappa(M-Y) - g Y \\
\frac{dZ}{dt} &= fq\kappa_\tau(M_\tau-Y_\tau)e^{-g\tau} - g Z \\
\frac{dX}{dt} &= h (H-X) - rX  \\ \\ \hline \\ 
\kappa &= c \frac{X(t)}{H} \\
h &= b fq \frac{Z(t)}{H} \\
\end{array}
\end{equation}

<center> 

![A diagram of the a version of the Ross-Macdonald model, using equations from Aron and May [@AronJL1982PopulationDynamics]](Figures/AronMay.png)

</center> 

These equations describe processes in three domains (Figure 2.1): 

+ adult mosquito ecology ($M$, and perhaps $V$); 

+ parasite infection dynamics in mosquito populations ($Y$ and $Z$); 

+ parasite infection dynamics in human populations ($X$). 

The equations describing parasite infections in mosquito populations also include the variable $M$, so the mosquito infection dynamics are coupled to the mosquito population dynamics.  The way we've written the equations, each compartment has an input term (*i.e.*, $\Lambda$, $\kappa$, or $h$) that depends on something else. We've passed $\Lambda$ as a parameter. For the infection dynamics, the terms $\kappa$ and $h$ couple two separate systems. For adult mosquito dynamics, emergence is passed to the model as a parameters. 

There are, of course, more compact ways of writing these equations. We have written the equations this way to emphasize a few things. First, the terms make it clear exactly how the equations in one domain are connected to another. Second, if we wanted to start *changing* some of the assumptions, these terms help to isolate the parts we might like to change. By writing the equations in this modularized form, we can start to understand how we might be able to write software that would allow us to represent mosquito infection dynamics with different systems of equations.  

The next step is to find solutions.  

*NOTE:* We don't introduce `exDE` or `MicroMoB` until [Modularity and Software].

## Solutions 

What does a **solution** to these equations look like? 

Solutions to these equations are values of the variables over time $\left( M(t), Y(t), Z(t), X(t) \right)$ that satisfy the system of four equations described above. We call these solutions *orbits.* To put it another way, if we took the derivatives of the orbits for any variable at any point in time using the basic definition $$\lim_{h\rightarrow 0} \frac{x(t+h)-x(t)}{h},$$ and then we used the values of the variables at time $t$ to compute $dM/dt$, $dY/dt$, $dZ/dt$, and $dX/dt$ (*i.e.*, using the formulas), we would get the same values.  

It is important that these orbits are unique: after specifying the *initial values* of the variables, there is one and only one set of orbits that solves the equations. When we solve the equations, we usually produce solutions from a starting point into a future, but the orbits are defined for all time -- $i.e.$ the process implies the existence of solutions far back into the past. These are deterministic equations, after all. 

As written, the equations do not define a *model.* Instead, the equations define a process or a **model family.** A model is something that *can* produce orbits. A model is defined only after assigning specific values to the parameters. Informally, we will often slip and use the "model" to describe a model family. It's easy to slip up, and sometimes we can get by with being sloppy, but we need to remember the distinction. When we say that the software is *modular,* we mean that it is easy to swap out one *model family* for another. 

To find solutions of equations we use an R software package called `deSolve.` Because of the delay for the EIP, these are called *delay differential equations,* which are handled using a function called `dede`. An important step in solving delay differential equations is a function `lagvalue()` that computes and returns the values of variables at a time lag, $\ell$. In these equations, the lag is set by the EIP, $\tau$, so we must evaluate 
`lagvalue(t-tau).`

In solving *ordinary differential equations,* we must pass initial conditions. To solve a delay differential equations with a maximum lag $\ell$, we must specify the initial conditions for the interval $[-\ell, t_0)$, where $t_0$ is the point in time when we start computing solutions.
In these equations, since the equation for $dZ/dt$ looks back $\tau$ units, we must specify values of $M(t)$, $Y(t)$, and $X(t)$ for all values of $t \in [-\tau, t_0)$. This forces an awkward choice, since we don't know the solutions backwards in time, but would need to know those solutions to use them. What is typically done -- and we've done it here -- is to specify a constant set of initial values and moving on. 

Doing this introduces a little *numerical slop.* By slop, we mean that these values are *not* what we would get if we ran the equations backwards in time. In these equations, it won't affect our analysis most of the time, so we're happy to acknowledge this little problem and find ways around it. It's a little thing, but we should never forget it, because we might find that it *is* affecting our analysis at some point.  

With `deSolve,` solving differential equations is not difficult -- it just involves following a few steps. In the following, we walk through these steps: 

+ Write a function that computes the derivatives;

+ Define initial conditions;

+ Define the values of the parameters;  

+ Define a mesh on time; 

+ Call a function that solves the equations, such as `dede` for delay differential equations.  

Many users will find that reading this code is like learning how to compute $\sqrt{2}$. If so, feel free to learn it once and then skip it. 

### Derivatives

The first step is to write down the equations to compute the derivatives. The solver expects a function with three required arguments (in this order): 

+ `t` is time 

+ `y` is the list of variables

+ `params` is a set of parameters 

The derivatives are computed and returned in the same order as 'y' in a `list`. To make code that is easy to read, we make `params` as a `list` with parameter names (see below), so that inside the function `with(params,{...})`, the parameter names are visible. 

```{r dAronMay}
dAronMay = function(t, y, params){with(params,{
 
  # Variables  
  if(t<=tau) ylag<-y0 else ylag <- lagvalue(t-tau)
  M=y[1]; M_tau = ylag[1]
  Y=y[2]; Y_tau = ylag[2]; 
  Z=y[3]; 
  X=y[4]; X_tau = ylag[4]
   
  # Terms 
  kappa = c*X/H; kappa_tau = c*X_tau/H
  h = b*f*q*Z/H 
   
  # Dynamics 
  dM = Lambda(t) - g*M
  dY = f*q*kappa*(M-Y) -g*Y
  dZ = f*q*kappa_tau*(M_tau-Y_tau)*exp(-g*tau) -g*Z
  dX = h*(H-X)-r*X
  
  return(list(c(dM, dY, dZ, dX)))
})} 
```

### Initial Values 

To run the model, we must supply initial values. If you were writing code yourself, it would be important to remember that the initial values and the return value for the derivatives must occur in the same order.   

A useful convention in {R} is to pass the initial values as a named list. Later, we can turn the outputs into a data frame, and then we can retrieve the variables by name. 

```{r}
y0= c(M=60, Y=0, Z=0, X=1)
```

The object `y0` is a named list -- the names are attached but invisible. 

```{r}
y0
```

When we turn it into a list, with `as.list,` the names are attached to the values: 

```{r}
as.list(y0)$M
```

If we use `with`, we create an environment where we can simply use the names: 

```{r}
with(as.list(y0), {
  M
})
```


### Parameter Values 

We pass the parameters as a list. It might seem like overkill, but we have written a function `makeParams()` that takes default values and generates a list. This makes it easy to generate a new set of parameter values with alternative values, and it also helps us to write and pass function $\Lambda(t)$ with parameters we like. By passing the parameter as a list, the parameter values are available to the function `dAronMay` when we use `with(params, {})`. 

Note that we have also attached the initial values of the variables as a parameter set, which are the return values for `lagvalue(t)` when `t<0`.   

```{r}
makeParams = function(y0, 
                      g=1/12, f=1/2.5, q=0.95,  
                      c=0.15,
                      b=0.55, r=1/200, H=1000,  
                      m=.05, ss=1,  
                      tau=10  
                      ){
  ss = min(1,max(0, ss))
  Lambda = function(t){m*H*(1 + ss*sin(2*pi*t/365))}
  return(list(y0=y0,g=g,f=f,q=q,c=c,
              H=H,m=m,tau=tau,b=b,r=r,Lambda=Lambda))
} 
params = makeParams(y0)
```

To make it absolutely clear, we are assuming: 

+ $g=1/12$: mosquitoes live about $12$ days, on average

+ $f=1/2.5$: mosquitoes feed every 2.5 days, on average 

+ $q=0.95$: the human fraction is 95\%; mosquitoes feed on humans 95\% of the time 

+ $c=0.15$: about 15\% of bites on infectious humans infect a mosquito 

+ $b=0.55$: about 55\% of bites by infective mosquitoes cause an infection

+ $r=1/200$: human infections last about $200$ days, on average 

+ $H=1000$: we're simulating transmission in a population of a thousand humans

+ $\tau=10$: the extrinsic incubation period is about 10 days 

+ For emergence, we tune the average value using $m$ and it is scaled to $H$: 

    - The parameter $m$ in the function above has been set to $0.05$ by default. 
    
    - The parameter $ss$ affects the amplitude of the fluctuations. We force it to take on values between 0 and 1. 
    
    - Emergence is modeled as a sinusoidal function with a yearly cycle. 
    
$$\Lambda(t) = m H \left(1 + \sin \left(\frac{2\pi t}{365}\right)\right)$$

### Solving 

We define a mesh over time -- the points in time when we would like to know the values of the variables: 

```{r}
tt = seq(0,5*365, by=5) 
```

This code solves the equations: 

```{r, message=F, warning=F}
require(deSolve)
yout <- dede(y=y0, times=tt, func=dAronMay, parms=params) 
```

### Visualizing 

```{r, echo=F}
plotTS_AronMay = function(yout){with(data.frame(yout),{
  par(mfrow = c(2,1))
  plot(time/365, M, type = "l", col = "blue", ylim = range(0,M), 
       xlab = "Time (in Years)", 
       ylab = expression(list(M(t), Y(t), Z(t))), 
       main = "Mosquitoes")
  lines(time/365, Y, col = "purple")
  lines(time/365, Z, col = "red")
  
  plot(time/365, X, ylim = c(0,1000), type = "l", 
       xlab = "Time (in Years)", 
       ylab = expression(X(t)), 
       main = "Humans")
})}
```

When we plot out the solutions, they look like this. 

```{r, fig.height= 6.5, fig.width = 5.5, echo=F}
plotTS_AronMay(yout)
```


\clearpage

## Steady States

Here, we analyze the system of equations in a narrow case when there is *no seasonality,* and the system reaches a *steady state.* To do so, we set the parameter `ss=1`, so that $\Lambda(t)$ is a constant; the resulting system is *autonomous.* We do this, in part, because the resulting system is easier to understand. We can develop intuition that can be applied (albeit with caution) to more complex systems. To be clear, we are dealing with this system: 

\begin{equation}
\begin{array}{rl}
\frac{dM}{dt} &= \Lambda - g M \\
\frac{dY}{dt} &= fq\kappa(M-Y) - g Y \\
\frac{dZ}{dt} &= fq\kappa_\tau(M_\tau-Y_\tau)e^{-g\tau} - g Z \\
\frac{dX}{dt} &= h (H-X) - rX  \\ \\ \hline \\ 
\kappa &= c \frac{X(t)}{H} \\
h &= b fq \frac{Z(t)}{H} \\
\end{array}
\end{equation}

As before, we have put the equations in their modularized form above, and the connecting terms below. 

The first thing to note is that $M$ affects $Y$ and $Z$, which affect $X$; but $M$ is not affected by $Y$ or $Z$. Mosquito population density is *exogenous* to malaria dynamics. 

### Mosquito Density 

We can thus treat it separately in the analysis:  

\begin{equation}
\frac{dM}{dt} = \Lambda - g M 
\end{equation}

Since emergence rates are steady, mosquito population density reaches a steady state when $dM/dt=0$, which occurs at: 

\begin{equation}
\bar M = \frac{\Lambda}{g} 
\end{equation}

### EIR 

Next, we note that at a steady state, the delayed values of variables and terms don't change, so from $dY/dt$, we get: 

\begin{equation}
g \bar Y = fq\kappa(\bar M- \bar Y) 
\end{equation}

If we substitute the formula for $\bar M$ and solve for $\bar Y$, we get: 

\begin{equation}
\bar Y = \frac{fq\kappa}{g + fq\kappa} \frac{\Lambda}{g}
\end{equation}

and since at the steady state, any variable at time $t+\tau$ is equal to its value at time $t$, we substitute the formula for $g \bar Y$ into $dZ/dt$ to get:  

\begin{equation}
g \bar Y e^{-g\tau} = g \bar Z
\end{equation}

Solving for $\bar Z$ we get: 

\begin{equation}
\bar Z =  \frac{f q \kappa}{g + fq \kappa} \frac{\Lambda}{g} e^{-g\tau} 
\end{equation}

At the steady state, $$\mbox{EIR} = fq \frac{\bar Z}{H}.$$ 

In field studies, the EIR is the product of the HBR and the sporozoite rate (SR). The sporozoite rate (SR, $z$) is given by: 

\begin{equation}
\bar z =  \frac{Z}{M} = \frac{f q \kappa}{g + fq \kappa} e^{-g\tau} 
\end{equation}

So we can understand the EIR as having two parts:

\begin{equation}
\mbox{EIR} = \mbox{HBR} \times  \mbox{SR} 
\end{equation}

or equivalently 

\begin{equation}
\mbox{EIR} = \frac{\textstyle{fq\Lambda}}{\textstyle{H}} \times \frac{\textstyle{f q \kappa}}{\textstyle{g + fq \kappa}} e^{-g\tau} 
\end{equation}

This formula for the SR (albeit with slightly different notation) was originally derived as part of the Ross-Macdonald model [@MacdonaldG1952Sporozoite; @ArmitageP1953]. Also, Smith and McKenzie (2004) have written a useful discussion of mosquito demography [@SmithDL2004_Statics]. 

### Vectorial Capacity 

Here, we pause to define a term that describes the number of human blood meals each mosquito would take over its whole life: 

$$S = \frac{fq}{g}.$$ 

Since $1/g$ is the mosquito lifespan in days, and $fq$ is the human blood feeding rate, $S$ is the number of human bloodmeals a mosquito would take over its lifespan. Intuitively, it makes sense that this *should* be what drives transmission, since it takes two human blood meals to transmit malaria parasites. 

If we rearrange the terms a bit, we can rewrite out the expression for the EIR: 

\begin{equation}
\mbox{EIR}(\kappa) = fq \frac{\bar Z}{H} = \frac{\Lambda}{H} S^2  e^{-g\tau} \frac{\kappa}{1 + S \kappa} 
\end{equation}

This formula for the EIR has two parts. We call the first part *vectorial capacity* ($V$): 

\begin{equation}
V = \frac{\Lambda}{H} S^2  e^{-g\tau} 
(\#eq:VCdefined)
\end{equation}

The second part is an expression that involves mainly $\kappa$. 

\begin{equation}
\frac{\kappa}{1 + S \kappa} 
(\#eq:EIR2ndpart)
\end{equation}

The relationship between VC and EIR at a steady state is a product: 

\begin{equation}
\mbox{EIR}(\kappa) = V \frac{\kappa}{1 + S \kappa} 
(\#eq:EIR2VC)
\end{equation}

Vectorial capacity describes the slope of the EIR when $\kappa$ is small: 

\begin{equation}
\left. d\frac{\mbox{EIR}(\kappa)}{d\kappa}\right|_{\kappa = 0} = V
(\#eq:VCisdEIR)
\end{equation}

We say that VC describes *potential* transmission, even if the parasites are absent. Another way to say the same thing is that when $\kappa$ is small, then:

\begin{equation}
\mbox{EIR}(\kappa) \approx V \kappa
\end{equation}

We can interpret vectorial capacity ($V$) in simple terms. It describes *the number of infective bites that would arise from all the mosquitoes biting a single human on a single day* but only *if all those mosquitoes became infected.*  Vectorial capacity tells the story of potential parasite transmission by mosquitoes in four steps, which highlights the fact that two human blood meals are required for the parasite to be transmitted and complete its life-cycle. 

\begin{equation}
\begin{array}{|c|c|c|c|c|c|c|}
\Lambda/H &  & S \kappa &  & e^{-g\tau} & & S \\
& \rightarrow &  & \rightarrow &  & \rightarrow &  \\
\mbox{Mosquito} & & \mbox{Parasite} & & \mbox{Mosquito} && \mbox{Parasite} \\
\mbox{Emerges} & & \mbox{Infects} & & \mbox{Survives} && \mbox{Infects} \\
 & & \mbox{Mosquito} & & \mbox{EIP} && \mbox{Human}
\end{array}
(\#eq:VCstory)
\end{equation}

As a reminder, while Eq. \@ref(eq:VCstory) includes $\kappa$, the formula for VC, in Eq. \@ref(eq:VCdefined), assumes that $\kappa=1$: the VC describes transmission as if humans were perfectly infectious. It was defined this way on purpose: it was meant to include mosquito parameters and exclude human factors. We can think of VC as defining something like a conditional expectation, a maximum, or (as we have already said) a measure of *potential transmission by mosquitoes* that is independent of human factors. 

While $\kappa$ (the numerator in Eq.\@ref(eq:EIR2ndpart) accounts for *most* of the difference between the EIR and the VC, the rest of the difference is due to the denominator in Eq. \@ref(eq:EIR2ndpart), $1+S\kappa$, which traces back to the formula from $dY/dt$, which assumes that mosquitoes are either infected or not. The denominator is a measure of saturation -- the fraction of mosquitoes that get *superinfected* with parasites. The main point here is that as $\kappa$ increases, saturation increases. If we set $S$ to the values in the previous plots, we can isolate the relationship: 

```{r, echo = F, fig=T, fig.cap="The effect (compare the solid and dashed black lines) and effect size of saturation (blue), graphically."}
par(mar= c(5,4,1,4))
S = with(params, f*q/g) 
mk=0.15
kappa = seq(0,.32, length.out=100)
plot(kappa, kappa/(1+S*kappa), type = "l", xlab=expression(kappa), ylim = c(0,mk), ylab = "")
lines(kappa, mk/(1+S*kappa), col = "blue")
segments(0,0,mk,mk, lty = 2)
axis(4, mk*c(0:5)/5, c(0:5)/5, col = "blue")
mtext(expression(kappa/(1+S*kappa)),2, 2.8)
mtext(expression(1/(1+S*kappa)), 4, 2.8, col = "blue")
axis(4, mk*c(0:5)/5, c(0:5)/5, col = "blue", col.ticks ="blue", col.axis="blue")
```

In these formulas, the measure of saturation is exactly $1+S\kappa$. We could rewrite the relationship between the EIR and VC in a way 
that tells us something about how we might be underestimating a parasite's reproductive success: 

$$\mbox{EIR}\times (1 + S \kappa) = V \kappa$$
which suggests that each infectious bite is passing along an excess $S\kappa$ bites. 


### Malaria Prevalence & Thresholds

We let $x$ denote infection prevalence: 

\begin{equation}
x = \frac{X}{H} 
\end{equation}

so $\kappa = c x$, and  

\begin{equation}
\frac{dx}{dt} = \frac{1}{H} \frac{dX}{dt} = h (1-x)-r x
\end{equation}

We can also define the basic reproductive number:

\begin{equation}
R_0 = \frac{bcV}{r}.
\end{equation}

It is the product of four terms: 

+ Vectorial capacity, $V$, the number of infective bites, per person, per day; 

+ The number of days a person would remain infectious, $1/r$; 

+ The fraction of infectious bites that would infect a human, $b$;  

+ The fraction of blood meals on infectious humans that would infect a mosquito, $c$  

After taking their product, we can interpret $R_0$ as a measure of the parasite's reproductive success after a single generation. It only depends on where we start counting. It could be one of the following:

+ the number of infected mosquitoes that would arise from a single infected mosquito; 

+ the number of infectious mosquitoes that would arise from a single infectious mosquito; 

+ the number of infected and infectious humans that would arise from a single infected and infectious human. 

Here, $R_0$ plays an important role in these equations if we start with $dX/dt$; then transform it to $dx/dt$; then replace $h$ with $bE$; then replace $\kappa$ with $cx$; then divide by $r$; and rearrange: 

\begin{equation}
\frac{1}{r} \frac{dx}{dt} = x \left[R_0 \left(\frac{1-x}{1 + cSx} \right)  - 1\right]  
\end{equation}

Since $x$ is the prevalence, it is always in the interval $[0,1]$. When $x$ is very close to $0$, then 

\begin{equation}
\frac{1-x}{1 + cSx} \lesssim 1. 
\end{equation}

and as $x$ grows very small: 

\begin{equation}
\lim_{x \rightarrow 0} \frac{1-x}{1 + cSx} = 1. 
\end{equation}

It follows that when $x$ is small, $dx/dt>0$ if and only if $R_0 > 1$. Depending on $R_0$, only one of two possibilities can hold:

+ either $R_0<1$, so that $x=0$ is the steady state; 

+ or $R_0 > 1$, and the steady state is: 

\begin{equation}
\bar x = \frac{R_0 -1}{R_0 + c S} 
\end{equation}

Since at the steady state, $\kappa = c \bar x$, we can plug this back into the formulas above to get $\bar Y$ and $\bar Z$. 

What we've learned about these equations is that if mosquito population densities are constant, then malaria reaches a steady state: if $R_0 >1$, then there is a positive endemic equilibrium, and if $R_0 < 1$, then malaria is absent from the system. The system is said to be stable -- in fact, is is globally asymptotically stable, which means that  all the orbits end up converging to the steady state. This statement has been proved many times in many papers, and since this book is focused on policy, we'll let others worry about proofs.


### Checking our Work 

An advantage of working in this environment is that we can check our work. One way we could solve these equations would be to run them for a very long time: 

We make a parameter set that defines the model: 

```{r}
y0=c(M=100, Y=10, Z=5, X=200)
paramsSteady = makeParams(y0, ss=0)
```


```{r, message=F, warning=F}
dede(y0, times=tt, func=dAronMay, parms=paramsSteady) -> yout
tail(yout, 1)[-1] -> eq1 
eq1
```

By plotting it out, we can check to see if we've run it for long enough:

```{r, fig.height= 6.5, fig.width = 5.5, echo=F}
plotTS_AronMay(yout)
```

We can check our algebra by computing the same quantities, and $R_0$ and other quantities we care about:  

```{r}
steadyStates_AronMay = function(params){with(params,{
  Lambda = m*H
  Meq = Lambda/g
  S = with(paramsSteady, f*q/g)
  V = m*S^2*exp(-g*tau)
  R0 = b*c*V/r
  x = ifelse(R0>1,(R0-1)/(R0+c*S), 0) 
  Xeq = x*H
  kappa = c*x
  Yeq = S*kappa/(1+S*kappa)*Meq 
  Zeq = Yeq*exp(-g*tau) 
  EIR = f*q*Zeq/H
  FoI = b*EIR
  aEIR = 365*EIR 
  aFoI = 365*FoI 
  extra=c(S=S, V=V, R0=R0, x=x, kappa=kappa, 
          EIR=EIR, FoI=FoI)
  annual =c(aEIR = aEIR, aFoI=aFoI)
  list(std=c(M=Meq, Y=Yeq, Z=Zeq, X=Xeq), 
       extra=signif(extra, 3),
       annual = signif(annual,3)) 
})}
steadyStates_AronMay(paramsSteady) -> eq2
eq2
```
Now, we can compare directly: 

```{r}
rbind(eq1=eq1, eq2=eq2$std)
```


## Stable Orbits 

If emergence rates vary seasonally, how much of the analysis that we did to understand *steady states* still holds?  Obviously, if conditions are changing seasonally, the model does not reach a steady state. In fact, after modification to suit the context, many of the same principles translate. The steady state analysis provides a good qualitative guide, but that the answers will look different. Here, we illustrate by solving systems to illustrate some basic points, which is easy enough. Analysis of the resulting dynamics can be quite difficult; it is covered in [Temporal Dynamics].   

```{r, echo=F, }
compareTS_AronMay = function(yout1, yout2){
  dta1 = data.frame(yout1) 
  dta2 = data.frame(yout2) 
  par(mfrow = c(2,2))
  with(dta1,plot(time/365, M, type = "l", col = "blue", 
       xlab = "Time (in Years)", 
       ylab = expression(M(t)), 
       main = "Mosquitoes"))
  with(dta2, lines(time/365, M, col = "blue", lty=2))
  yrng = range(dta1$Y, dta2$Y)
  with(dta1,plot(time/365, Y, type = "l", ylim = yrng, col = "purple", 
       xlab = "Time (in Years)", 
       ylab = expression(Y(t)), 
       main = "Infected Mosquitoes"))
  with(dta2, lines(time/365, Y, col = "purple", lty=2))
  zrng = range(dta1$Z, dta2$Z)
  with(dta1,plot(time/365, Z, type = "l", ylim = zrng, col = "red", 
       xlab = "Time (in Years)", 
       ylab = expression(Z(t)), 
       main = "Infectious Mosquitoes"))
  with(dta2, lines(time/365, Z, col = "red", lty=2))
  xrng = range(dta1$X, dta2$X)
  with(dta1,plot(time/365, X, ylim = xrng, type = "l", 
       xlab = "Time (in Years)", 
       ylab = expression(X(t)), 
       main = "Infected Humans"))
  with(dta2, lines(time/365, X, lty=2))
}
```

### Thresholds

There is a threshold condition $R_0>1$ that determines whether malaria is endemic, but the formula for $R_0$ depends on the form of $\Lambda(t)$. If we set $R_0=1$, we can show that the threshold for persistence in a seasonal environment is $R_0 > \sigma > 1$ (see Figure 1.1). The math to compute threshold conditions in seasonal environments is in [Temporal Dynamics]. 


```{r, echo=F}
makeParamsThresh = function(y0, 
                      g=1/12, f=1/2.5, q=0.95,  
                      c=0.15,
                      b=0.55, r=1/200, H=1000,  
                      ss=1, R0=1, 
                      tau=10  
                      ){
  S = f*q/g 
  m = R0*r/(b*c*S^2*exp(-g*tau))
  ss = min(1,max(0, ss))
return(list(y0=y0,g=g,f=f,q=q,c=c,H=H,tau=tau,b=b,r=r,
  Lambda = function(t){m*H*(1 + ss*sin(2*pi*t/365))})) 
} 
``` 


```{r, echo=F, fig.cap = "Here, we set $R_0= 1.02$ for the model with constant emergence, and we show that malaria persists. For the same parameters and for the same *average* emergence rate, malaria declines with seasonality."}
yI = c(M=2, Y=0, Z=0, X=10)
tt = seq(0, 7300, by=5)

paramsA = makeParamsThresh(yI, R0=1.02, ss=0)
paramsB = makeParamsThresh(yI, R0=1.02)
youtA <- dede(y=yI, times=tt, func=dAronMay, parms=paramsA) 
youtB <- dede(y=yI, times=tt, func=dAronMay, parms=paramsB) 

compareTS_AronMay(youtB, youtA)
```


\clearpage

### Orbits  

If $R_0 >1$, then all orbits converge to a set of *stable orbits* (See Figure 1.1). If $\Lambda(t)$ has an annual cycle, then after the orbits converge: 

    - $M(t+365) = M(t)$; 
    - $Y(t+365) = Y(t)$ and $Z(t+365) = Z(t)$; 
    - $X(t+365) = X(t)$. 


```{r, echo=F, fig.cap = "With different initial values, the orbits converge and eventually lie on top of one another."}
y1 = c(M=1100, Y=100, Z=100, X=100)
tt = seq(0, 830, by =1)
params1 = makeParams(y1)
yout0 <- dede(y=y0, times=tt, func=dAronMay, parms=params) 
yout1 <- dede(y=y1, times=tt, func=dAronMay, parms=params1) 
compareTS_AronMay(yout0, yout1)
```

\clearpage

### Average Dynamics 

If $R_0>1$ and malaria is endemic, the *average* prevalence of malaria infection is variable in a seasonal environment. While the prevalence is higher at the peak, the average for the whole year tends to be lower.  

```{r compareTS, echo=F, fig.cap = "Here, we set $R_0= 2$ for the model with constant emergence, and we show that the prevalence of malaria is similar in the seasonal environment, but it's higher as transmission peaks, lower in the off-season, and lower overall."}
yI = c(M=20, Y=5, Z=1, X=100)
tt = seq(0, 7300, by=5)

paramsC = makeParamsThresh(yI, R0=2, ss=0)
paramsD = makeParamsThresh(yI, R0=2)
youtC <- dede(y=yI, times=tt, func=dAronMay, parms=paramsC) 
youtD <- dede(y=yI, times=tt, func=dAronMay, parms=paramsD) 

compareTS_AronMay(youtD, youtC)
```

## Discussion

If we were simply learning the math, ...

When we analyze these equations to determine their *stability* and to identify *threshold conditions,* we focus on threshold conditions and the behavior of these systems when malaria is rare. In most places, malaria is endemic so we need to be concerned about malaria immunity and its effects on transmission; malaria is under some level of control; and because of weather and other factors, the baseline conditions change from year to year. 



