---
output:
  html_document: default
  pdf_document: default
---

# Stage of Infection Models

This chapter introduces **Stage of Infection** models. 

## SoI as an ODE Model

### Simple SoI 

One way to model infections is to break them up into "stages."

+ $U$ is the density of uninfected individuals.

+ $E$ is exposed but not yet infected; the duration of latency is $1/\nu$ days. 

+ $X_i$ is infected and infectious in stage $i=1 \ldots N$, and the duration in stage $i$ is $(\nu_i + r_i)^{-1}$, with a proportion $\nu_i /(\nu_i + r_i)$ proceeding onto the next stage, while the rest simply clear. (Obviously, $\nu_N = 0$). 

\begin{equation}\begin{array}{rl}
\dot U & = - h U + \sum_N r_N X_N \\ 
\dot E &= h U - \nu_0 E \\ 
\dot X_1 & = \nu_0 E - (\nu_1 + r_1) X_1 \\ 
\dot X_i & = \nu_{i-1} X_{i-1} - (\nu_i + r_i) X_i \\ 
\end{array}\end{equation}

+ The average duration of an infection (post latency) is the sum of exponentials (which is $\Gamma$-like) with an average duration 
$$\sum_i \prod_{j<i} \frac{\nu_j}{\nu_j + r_j} (r_i + \nu_i)^{-1}$$

### Stage of Infection, with Superinfection 

We add the variable $E_i$, people who are infected in stage $i$ and exposed. 

\begin{equation}\begin{array}{rl}
\dot U & = - h U + \sum_N r_N X_N \\ 
\dot E_0 &= h U - \nu_0 E_0 \\ 
\dot E_i &= h X_i - \nu_0 E_i \\ 
\dot X_1 & = \sum_i \nu_i E_i - (\nu_1 + r_1) X_1 \\ 
\dot X_i & = \nu_{i-1} X_{i-1} - (\nu_i + r_i + h) X_i \\ 
\end{array}\end{equation}

+ Note that the average duration of an infection is now implicitly related to $h$.   

## SoI as a Tensor Model

### 4SoI+S-PIT*

The following document describes a model with the following properties:

+ It is a stage-structured model with 4 stages of infection (4SoI)

+ It allows for overwriting superinfection (+S)

+ It uses PIT variables for immunity (-PIT)

+ It also considers antimalarial drug use and chemoprotection (-Rx)

+ This is version 1.0

## Stage Structured Infections with Overwriting Superinfection 

+ The time step is $10$ days ($\tau=10$). 

+ Infections are stage-structured, including a latent class $E_0$. The number of people in a population with active blood stage infection in stage $i$ is $X_1, X_2, \ldots X_N$. The clearance rate in the $i^{th}$ stage is denoted $r_i$; the rate of advancement to the $i^{th}$ stage is $\nu_i$.

+ Superinfection can occur at any stage; superinfected individuals enter an exposed class $E_i$ corresponding to their infection stage, and then re-enter stage 1, either $X_1$ or $E_1$. 

+ Fever occurs in a fraction $\eta$ of the population in stage 1. 

+ A fraction $\rho_t$ of febrile cases are treated. Case management is followed by a period of chemoprotection lasting two time periods, or 20 days. The model thus includes a variable describing a single period of chemoprotection, $P$, after which individuals are forced to re-enter the uninfected class.

+ The total population is constant $H = U + \sum_i E_i + \sum_i X_i + P$. 

The discrete-time version of this model can be represented as a set of difference equations, with constants defined as follows (dropping the subscript for time): 

+ $D = 1-e^{-\delta \cdot10}$ and $D' = 1-D$

+ $T = \eta \rho + (1-\eta) (1-e^{-\delta \cdot10})$ and $T' = 1-T$  

+ $A_t' = 1-A_t$

+ $S_i = e^{-(r_i + \nu_i)10}$

+ $Q_i = r_i /(r_i + \nu_i) (1-S_i)$

+ $V_i = \nu_i /(r_i + \nu_i) (1-S_i)$

\begin{equation}\begin{array}{rl}
P_{1,t+1} &= D \left( U_{i,t} + \sum_i X_{i,t} \right) + T \sum_i E_i \\
%
P_{2,t+1} &= P_{1,t} \\ 
%
U_{t+1}   &= A_t' D' P_{2,t} + A_t' D' U_t + \sum_i Q_i X_{i,t} \\ 
%
X_{1,t+1} &= D' \left(A_t' P_{J,t} + A_t' S_1 X_{1,t} + T' A_t' E_1 + \sum_{i \neq 1} E_{i,t} \right) \\
X_{4,t+1} &= A_t' D' \left( S_i X_{i,t} + V_{3} X_{4,t} \right) \\
%
E_{0,t+1} &= D' A_t \left( U_t + \sum_i Q_i X_{i,t} \right)  \\
E_{1,t+1} &= D' A_t S_1 X_{1,t} + T' A_t E_1  \\ 
E_{i,t+1} &= D' A_t X_{i,t} \\
%
\end{array}\end{equation}


In vector and matrix notation, the state space is $\vec Y_t = \left<P_1,P_2,U,X_1,\ldots,X_N, E_0,\ldots,E_N \right>,$ and for $N=4$, and $J=1$ we get ${\cal P}=$

\begin{equation*} \left[ 
\begin{array}{c|cccccccccccc}
& P_1 & P_2 & U & X_1 & X_2 & X_3 & X_4 & E_0 & E_1 & E_2 & E_3 & E_4\\ \hline 
P_1 & 0 & 0 & D & D & D & D & D & T & T & T & T & T  \\
P_2 & 1 & 0 & D & D & D & D & D & T & T & T & T & T  \\
U & 0 & D' A' & D'A' & D'A'Q_1 & D'A'Q_2 & D'A'Q_3 & D'A'Q_4 & 0 & 0 & 0 & 0  & 0\\
X_1 & 0 & 0 & 0 & D'A'S_1 & 0 & 0 & 0 & T' & T'A' & T' & T' & T'  \\
X_2 & 0 & 0 & 0 & D'A'V_1 & D'A'S_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
X_3 & 0 & 0 & 0 & 0 & D'A'V_2 & D'A'S_3 & 0 & 0 & 0 & 0& 0 & 0   \\
X_4 & 0 & 0 & 0 & 0 & 0 & D'A'V_3 & D'A'S_4 & 0 & 0 & 0 & 0 & 0 \\
E_0 & 0 & 0 & D'A & D'A Q_1 & D'A Q_2 & D'A Q_3 & D'A Q_4 & 0 & 0 & 0 & 0 & 0 \\
E_1 & 0 & 0 & 0 & D'A S_1 & 0 & 0 & 0 & 0 & T' A & 0 & 0 & 0 \\
E_2 & 0 & 0 & 0 & D'A V_1 & D'A S_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
E_3 & 0 & 0 & 0 & 0 & D'A V_2 & D'A S_3 & 0 & 0 & 0 & 0 & 0 & 0 \\
E_4 & 0 & 0 & 0 & 0 & 0 & D'A V_3 & D'A S_4 &  0 & 0 & 0 & 0  & 0\\
\end{array}\right] 
\end{equation*}

which we can write as $B=AW+V$, where $V=$

\begin{equation*} \left[ 
\begin{array}{c|ccccccccccc}
& P & U & X_1 & X_2 & X_3 & X_4 & E_0 & E_1 & E_2 & E_3 & E_4\\ \hline 
P & 0 & D & D & D & D & D & T & T & T & T & T  \\
U & 1 & D' & D'Q_1 & D'Q_2 & D'Q_3 & D'Q_4 & 0 & 0 & 0 & 0  & 0\\
X_1 & 0 & 0 & D'S_1 & 0 & 0 & 0 & T' & T' & T' & T' & T'  \\
X_2 & 0 & 0 & D'V_1 & D'S_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
X_3 & 0 & 0 & 0 & D'V_2 & D'S_3 & 0 & 0 & 0 & 0& 0 & 0   \\
X_4 & 0 & 0 & 0 & 0 & D'V_3 & D'S_4 & 0 & 0 & 0 & 0 & 0 \\
E_0 & 0 & 0 & 0  & 0  & 0  & 0  & 0 & 0 & 0 & 0 & 0 \\
E_1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 &  0 & 0 & 0 & 0 \\
E_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
E_3 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
E_4 & 0 & 0 & 0 & 0 & 0 & 0 &  0 & 0 & 0 & 0  & 0\\
\end{array}\right] 
\end{equation*}

and $W=$

\begin{equation*} \left[ 
\begin{array}{c|ccccccccccc}
& P & U & X_1 & X_2 & X_3 & X_4 & E_0 & E_1 & E_2 & E_3 & E_4\\ \hline 
P & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
U & 0 & -D' & -D'Q_1 & -D'Q_2 & -D'Q_3 & -D'Q_4 & 0 & 0 & 0 & 0  & 0\\
X_1 & 0 & 0 & -D'S_1 & 0 & 0 & 0 & 0 & -T' & 0 & 0 & 0  \\
X_2 & 0 & 0 & -D'V_1 & -D'S_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
X_3 & 0 & 0 & 0 & -D'V_2 & -D'S_3 & 0 & 0 & 0 & 0& 0 & 0   \\
X_4 & 0 & 0 & 0 & 0 & -D'V_3 & -D'S_4 & 0 & 0 & 0 & 0 & 0 \\
E_0 & 0 & D' & D' Q_1 & D' Q_2 & D' Q_3 & D' Q_4 & 0 & 0 & 0 & 0 & 0 \\
E_1 & 0 & 0 & D' S_1 & 0 & 0 & 0 & 0 & T'  & 0 & 0 & 0 \\
E_2 & 0 & 0 & D' V_1 & D' S_2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
E_3 & 0 & 0 & 0 & D' V_2 & D' S_3 & 0 & 0 & 0 & 0 & 0 & 0 \\
E_4 & 0 & 0 & 0 & 0 & D' V_3 & D' S_4 &  0 & 0 & 0 & 0  & 0\\
\end{array}\right] 
\end{equation*}

#### RAMP Implementation

Since $W$ and $V$ depend on $\rho_t$ and $\delta_t$, we will need functions that generate these matrices for each time step: 

```{r makeParams.stage}
makeParams.stage = function(nu=1/80, r=1/200, eta=1/365, delta=1/730, N=4, T=10){
  Xix = 1 + 1:(N+1)
  Eix = 2 + N + 1:(N+1)
  E1 = 4+N
  X1 = 3
  dd = exp(-delta*T)  
  if(length(r) ==1) r=rep(r,N)
  if(length(nu)==1) nu=c(rep(nu,N-1),0)
  P = list(r=r,nu=nu,eta=eta,N=N,T=T,Xix=Xix,Eix=Eix,E1=E1,X1=X1,dd=dd)
  class(P) <- "stage"
  P
}

P = makeParams.stage(nu=c(1/20,1/80, 1/320,0))
```

```{r makeX.stage}
makeX.stage = function(P){with(P,{
  # stay
  s = exp(-(r+nu)*10)
  X = diag(s)
  # clear
  X = rbind(U=r/(r+nu)*(1-s), X)
  # advance stage
  xx = cbind(2:N+1, 2:N-1)
  v = (nu/(r+nu)*(1-s))[-N]
  X[xx] = v
  # unexposed
  X = cbind(U=0,X)
  X[1,1]=1
  #if(colSums(X)!=1) print("ERROR")
  return(X)
})}
#X = makeX.stage(P)
```

```{r makeV.stage}
makeV.stage = function(rho, delta, P){with(P,{
  X = makeX.stage(P)
  E=0*X
  E[2,] = 1
  
  V=matrix(0,2*N+3, 2*N+3)
  V[  1,Xix] = 1-dd
  V[Xix,Xix] = dd*X
  V[  1,Eix] = eta*rho
  V[  3,Eix] = 1-eta*rho
  V[2,1] = 1

  V
})}
V = makeV.stage(.2, 1/730, P)
```

```{r makeW.stage}
makeW.stage = function(rho, delta, P, V){with(P,{
  X = makeX.stage(P)
  E=0*X
  
  W=matrix(0,2*N+3, 2*N+3)
  W[Xix,Xix] = -dd*X
  W[Eix,Xix] = dd*X
  W[X1,E1] = -(1-rho*eta)
  W[E1,E1] = 1-rho*eta
  W
})}
W = makeW.stage(.2, 1/730, P, V)
```

```{r makeB.stage}
makeB.stage = function(A, rho, delta, P){with(P,{
  X = makeX.stage(P)
  E=0*X
  E[2,] = 1
  
  V=matrix(0,2*N+3, 2*N+3)
  V[  1,Xix] = 1-dd
  V[Xix,Xix] = dd*X
  V[  1,Eix] = eta*rho
  V[  3,Eix] = 1-eta*rho
  V[2,1] = 1

  V
})}
B = makeB.stage(.1, .2, 1/730, P)
```


## Detection 