
```{r}
require(Matrix)
```

## Age of Infection

The basic age of infection works as a shift operator. Let $\tau$ denote the time step. Let $s_i$ be the fraction that remains infected at the $i^{th}$ time step following the age of infection, $T_i$ the fraction that are treated and cured, and $q_i$ the fraction that clears its infection. Let $P$ denote the duration of the population that is in the $i^{th}$ period of chemoprotection.


The probability transition matrix is

\begin{equation} \left[ \begin{array} {c|ccccccccc|cccc}
    & U & X_0 & X_1 & X_2 & \ldots & X_{n-3} & X_{n-2} & X_{n-1} & X_n & P_1 & P_2 & \ldots & P_n \\ \hline
U   & 1-\alpha_t & q_0 & q_1 & q_2 & \ldots & q_{n-3} & q_{n-2} & q_{n-1} & q_n &  &  & & 1\\ 
X_0 & \alpha_t & . & &  &  & &  &  & \\ 
X_1 &   & s_0 & . &  &  & &  &  &\\ 
X_2 &  & & s_1 & . &  &  & &  &  &\\   
\vdots &  &  &  &  & \ddots &  &  & &  \\  
X_{n-3} &  &  & &  &  & . &  & &  \\  
X_{n-2} &  &  & &  &  & s_{n-3} & . & &  \\  
X_{n-1} &  &  & &  & &  & s_{n-2} & . &  \\  
X_n &  &  & &  &  &  &  & s_{n-1} & s_n \\  \hline
P_1   &  0 & T_0 & T_1 & T_2 & \ldots & T_{n-3} & T_{n-2} & T_{n-1} & T_n  & . &  &  & \\ 
P_2 &  &  &  &  &  &  & &  &  & 1 & . & &  \\  
P_3 &  &  &  &  &  &  & &  &  &  & 1 & . &  \\  
\vdots &  &  &  &  &  &  & &  &  &  &  & \ddots &  \\  
P_n  &  &  &  &  &  &  &  & &  &  &  &  & .\\  
\end{array} \right]  \end{equation}
```{r}

```

For the 10-day model, we let $X_0$ denote the incubation period.

```{r constructP_AoI}

makeParams_AoIten = function(
  nX = 21, 
  nP = 3,
  nr = 3,
  rho1 = 0.3,
  rho2 = 0.05,
  r = 1/200,
  pwr = 3,
  tau = 10
){list(nX=nX,nP=nP,nr=nr,r=r, rho1=rho1,rho2=rho2,pwr=pwr,tau=tau)} 

constructP_AoIten = function(alpha, par){ with(par,{
  eta = c(0,0, rho1, rho2*(c(nX:4)/nX)^pwr,0)
  s = c(alpha, rep(1, nr), rep(exp(-tau*r), nX-nr))*(1-eta)
  s = c(s,rep(1, nP))
  P = t(bandSparse(nX+nP+1, k=1)*s) 
  P[nX+2,1+0:nX] = eta
  P[nX+1, nX+1] = exp(-tau*r)
  P[1,] = 1-colSums(P)
  return(P)
})}
```

```{r demoP}
par = makeParams_AoIten()
constructP_AoIten(.01, par) -> Paoi
colSums(Paoi)
```
```{r X0}
vecX0 = with(par,matrix(c(1, rep(0, nX+nP))), 1, nX+nP+1) 
```

```{r make alpha}
#alpha = rbeta(146, 10*c(1+sin(2*pi*c(1:146)/36.5)), 100)/100
alpha = (1+sin(2*pi*c(1:146)/36.5))*0.01
plot(alpha, type="l")
```

```{r sim}
X = vecX0
for(a in alpha){
  X = constructP_AoIten(a,par) %*% X 
}
Xt=X
for(a in alpha){
  X = constructP_AoIten(a,par) %*% X 
  Xt = cbind(Xt, X)
}
```

```{r calD}
calDlm = with(par,matrix(c(0,0,c(nX:2)/(nX-1), rep(0,nP)),1,nX+nP+1))
calDtrue = with(par,matrix(c(0,0,rep(1, nX-1), rep(0,nP)),1,nX+nP+1))
```


```{r plot PR}
xlm = calDlm %*% as.matrix(Xt)
xtrue = calDtrue %*% as.matrix(Xt)
plot(0:146, xtrue, type = "l", ylim =range(0,xtrue))
lines(0:146, xlm, col = "red")
lines(1:146, alpha, col = "blue") 
```

```{r plot Sensitivity}
plot(0:146, xlm/xtrue, type ="l")
```

```{r plot Phase}
plot(alpha, c(xlm/xtrue)[-1], type = "l") 
```

