---
output:
  html_document: default
  pdf_document: default
---

# Tracking Variables

Suppose we have a variable rate of exposure, and we want to formulate models in which recent exposure affects the epidemiology of disease. We construct the following two time series tracking exposure:

```{r alpha, echo=FALSE}
alphaT_f = function(tmax, a=2, b=50){ 
  tt = 1:tmax
  rbeta(tmax, a*(1+sin(2*pi*tt/36.5)), b)
}
TT = 200
stepT = c(rep(1,TT), rep(0,TT))
alphaT = alphaT_f(2*TT)
ttTT = 0:(2*TT)*10/365
plot(ttTT[-401], alphaT, type = "l", ylab = expression (alpha[t]), xlab = "Time (Years)")
#lines(ttTT[-401],mean(alphaT)*stepT, col = "blue")
```

## Population Immune Tracking: Exposure Tracking

In a population stratified by age, cumulative exposure will vary by age. Let $I_{a,t}$ denote an immune tracking variable that tracks exposure for people of age $a$ at time $t$. It is updated by the following recursive relationship:

\begin{equation}
I_{a,t+1} = \alpha_t + (1-\mu) I_{a,t}
\end{equation}

Note that for $\mu=0$, the equation simply tracks cumulative exposure. This version returns only the last value: 
  
```{r basePIT.0}
basePIT.0 = function(alpha, mu=0, vecV = matrix(0,1,1), calO=matrix(1,1,1), N=1, w=1){
  ###################################################################
  # If alpha is a vector (a time series), then N is its length.
  # If alpha is a scalar, then N is passed as an optional argument.
  # Default values update one time step for alpha and mu.
  ###################################################################
  if(length(alpha)==1){
    alpha = rep(alpha, N)
  } else {
    N = length(alpha) 
  }
  ####################################
  # alpha is the "attack" rate
  # mu describes the "decay"
  # w describes biting weights by age
  # calO is an age operator
  ####################################
  for(t in 1:N){
    vecV = ((1-mu)*vecV + w*alpha[t])%*%calO
  } 
  return(vecV)
}
```

This version returns the whole time series (including the initial value):
```{r basePIT.TS}
basePIT.TS = function(alpha, mu=0, vecV = matrix(0,1,1), calO=matrix(1,1,1), N=1, w=1){
  ###################################################################
  # If alpha is a vector (a time series), then N is its length.
  # If alpha is a scalar, then N is passed as an optional argument.
  # Default values update one time step for alpha and mu.
  ###################################################################
  if(length(alpha)==1){
    alpha = rep(alpha, N)
  } else {
    N = length(alpha) 
  }
  ####################################
  # alpha is the "attack" rate
  # mu describes the "decay"
  # w describes biting weights by age
  # calO is an age operator
  ####################################
  matV = vecV
  for(t in 1:N){
    vecV = ((1-mu)*vecV + w*alpha[t])%*%calO
    matV = cbind(matV, vecV)
  } 
  return(matV)
}
```

We write a function `plotPITxAge` that plots PIT variables by age. 
```{r plotPITxAge}
plotPITxAge = function(ages, vecV, lines=FALSE, clr=NULL, d2y=F){
  clr <- ifelse(is.null(clr),"black", clr)
  if(d2y == TRUE) ages <- ages/365
  if(lines==FALSE){
    plot(ages, vecV, type = "l", xlab = "Age (Years)", ylab = "PIT", col = clr)
  } else { 
    lines(ages, vecV, col = clr)
  }
}
```

We use the age operator, ${\cal O$} and the `1066` age structure (from the `AgeOperators.R` in the the `Operational-Model-Library`). The following plots cumulative exposure by age. The age structure here tracks populations for 85 years. Black shows 85 years of exposure. Blue shows the results for 10,000 days ($\approx$27.4 years) and red for 1,000 days ($\approx$2.7 years). Starting from no exposure, we see the relationship saturate.  This is a cautionary tale to take care to set the initial conditions for tracking variables $V_0$ with great care. 
```{r eg-baseXage}
source ("./Operational-Model-Library/AgeOperators.R")
calO = calO_1066()
ages = ages_1066()
vecV = matrix(0,1,66)

plotPITxAge(ages, basePIT.0(0.1, mu=0, vecV, calO, N=3000))
plotPITxAge(ages, basePIT.0(0.1, mu=0, vecV, calO, N=1000), T, "blue")
plotPITxAge(ages, basePIT.0(0.1, mu=0, vecV, calO, N=100), T,"red")
```


We plot another function, `plotPITxTime` that plots PIT variables over time.

```{r plotPITxTime}
plotPITxTime = function(tm, vecV, lines=FALSE, clr=NULL, d2y=F){
  clr <- ifelse(is.null(clr),"black", clr)
  if(d2y == TRUE) tm <- tm/365
  if(lines==FALSE){
    plot(tm, vecV, type = "l", xlab = "Time (Years)", ylab = "PIT", col = clr)
  } else { 
    lines(tm, vecV, col = clr)
  }
}
```

The following shows how the PIT variables behave with respect to time. Exposure is plotted in red. Other colors track exposure with decay:
```{r}
vecV0 = 0
plotPITxTime(ttTT, basePIT.TS(alphaT, mu=0.05, vecV0)->V05)
plotPITxTime(ttTT, basePIT.TS(alphaT, mu=0.1, vecV0)->V10, T, "blue")
plotPITxTime(ttTT, basePIT.TS(alphaT, mu=0.2, vecV0)->V20, T,"darkred")
plotPITxTime(ttTT, basePIT.TS(alphaT, mu=0.3, vecV0)->V30, T,"darkgreen")
plotPITxTime(ttTT, basePIT.TS(alphaT, mu=0.5, vecV0)->V50, T,"purple")
lines(ttTT[-401], alphaT, col = "red")
```

This shows how the tracking variables work for constant exposure:
```{r}
vecV0 = 0
plotPITxTime(ttTT, basePIT.TS(stepT, mu=0.05, vecV0)->V05)
plotPITxTime(ttTT, basePIT.TS(stepT, mu=0.1, vecV0)->V10, T, "blue")
plotPITxTime(ttTT, basePIT.TS(stepT, mu=0.2, vecV0)->V20, T,"darkred")
plotPITxTime(ttTT, basePIT.TS(stepT, mu=0.3, vecV0)->V30, T,"darkgreen")
plotPITxTime(ttTT, basePIT.TS(stepT, mu=0.5, vecV0)->V50, T,"purple")
```
Isolating the decay:
```{r}
ix = 200:300
tix = c(ix-200)/36.5
plot(tix, V05[ix]/max(V05), type = "l", xlab = "Time (Years)", ylab = "Relative to Max")
lines(tix,V10[ix]/max(V10), col = "blue")
lines(tix,V20[ix]/max(V20), col = "darkred")
lines(tix,V30[ix]/max(V30), col = "green")
lines(tix,V50[ix]/max(V50), col = "purple")
```

## Cumulative Exposure

Let $W_t$ denote cumulative exposure, measured by the number of malaria attacks: 

\begin{equation}
W_{t+1} = W_{t} + \alpha_t
\end{equation}

Another measure is the cumulative number of infection days, weighted in the same way as detection. One application for this is that it is a measure of the total number of new  *var* genes that have been seen, where $c_\tau$ is the average number experienced per $\tau$-day time period. 

\begin{equation}
W_{t+1} = W_{t} + c_\tau x_t
\end{equation}

Alternatively, we can weight infection stages or ages in some other way, by their immunogenicity, $\cal I$: 

\begin{equation}
W_{t+1} = W_{t} + {\cal I}^T \vec X_t
\end{equation}

Two generic classes of counters are the following: 
\begin{equation}
W_{t+1} = p W_{t} + q {\cal I}^T \vec X_t
\end{equation}

and 

\begin{equation}
W_{t+1} = p W_{t} + q \alpha_t - s W_t^2
\end{equation}

```{r}
tau = c(1:365)*10
alpha = (1+sin(1+2*pi*tau/365))/4

makeW = function(p,q,alpha,s=0){
  Wl = W = 0
  for (i in 1:length(alpha)){
    W = p*W + q*alpha[i] - s*W^2
    Wl = c(Wl, W)
  }
  Wl[-1]
}

W = makeW(1, .1, alpha, s=.01)
plot(tau/365, W, type ="l")
lines(tau/365, alpha*max(W), col = "red")

```

```{r}

```

